Single-cell virus sequencing of influenza infections that trigger innate immunity

### 2019-03-18 paper/paper.tex
revise Intro and Results

\newfloat{suppfile}{thp}{lofsupfile}
\floatname{suppfile}{\bf FILE}
\renewcommand{\thesuppfile}{\bf S\arabic{suppfile}}

\newfloat{suppfig}{thp}{lofsupfile}
\floatname{suppfig}{\bf FIG}
\renewcommand{\thesuppfig}{\bf S\arabic{suppfig}}

@@

@@

Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single}.
Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

It is unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\cite{bhushal2017cell}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Because influenza has a high mutation rate, individual virions often have genetic defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{velthuis2018mini, du2018genome, killip2017single, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques primarily measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, zanini2018single, zanini2018virus, steuerman2018dissection, saikia2018simultaneous, oneal2018west}.
None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation.

Here we develop a new approach measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
To do this, we first enrich for rare IFN+ cells, and then perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
We obtain transcriptomes and sequences of all expressed viral genes in 150 infected cells, 40 of which express IFN.
Two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
This viral diversity is a major contributor to cell-to-cell heterogeneity, with cells infected by unmutated virions having a tighter distribution of viral transcriptional burden.
We identify several types of viral defects that increase IFN induction.
However, viral genetic variation does not fully explain the heterogeneity, and even unmutated virions sometimes induce IFN.
Therefore, viral diversity is an important but not exclusive cause of cell-to-cell heterogeneity during influenza infection.

@@

\subsection{A system to identify and enrich rare IFN+ cells}

@@

A challenge in studying IFN induction by influenza virus is the rareness with which infection triggers its expression.
Studies using reporters~\cite{killip2017single} and our own previous single-cell transcriptomics~\cite{russell2018extreme} have found that IFN induction is rare in cell lines infected with influenza.
IFN expression also appears to be rare among cells in influenza-infected mice~\cite{kallfass2013visualizing}, a conclusion supported by our re-analysis of single-cell transcriptomics of influenza-infected mice by Steuerman et al~\cite{steuerman2018dissection} (\SUPPFIG{mice}).

To identify and enrich rare IFN+ cells, we created A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the IFN reporter can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by infection with saturating amounts of a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that in our prior single-cell transcriptomics of influenza-infected cells.

\subsection{Combined transcriptomics and virus-sequencing of single infected cells}

@@

\caption{
Approach for combined transcriptomic and viral sequencing of single influenza-infected cells that express IFN.
(A)
IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
IFN+ cells are enriched by MACS (\SUPPFIG{MACS}), and pooled with non-enriched cells and uninfected canine cells that serve as an internal control for multiplets and mRNA leakage.
(B)
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
(C)
Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
(D)
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are enriched and sequenced by PacBio).
(E)
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

@@

To determine if mutations or other features of the infecting virions contribute to the heterogeneous outcome of infection, we developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide an important control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in about a quarter of cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter (\SUPPFIG{MACS}).
To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

We processed the cells on a commercially available platform~\cite{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
Because influenza virus mRNAs are polyadenylated~\cite{robertson1981polyadenylation}, this process appends cell barcodes to both cellular and viral mRNAs.
Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\cite{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\cite{robertson1981polyadenylation} in each of the eight viral gene segments.

We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs~\cite{travers2010flexible}).
These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).

\subsection{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells}

@@

\caption{
Single-cell transcriptomics of IFN-enriched influenza-infected cells.
(A) 
Number of cells for which transcriptomes were obtained.
From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are actually multiplets derived from multiple cells. 
(B) The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
(C) Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA in the human cells that are called as infected.
(D) Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
The majority of cells express all eight gene segments, but a substantial minority fail to express at least one gene.
\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
(E) Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell that is derived from each gene.
(F) Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
(G) Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line are classified as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
Many cells that do not express IFN still express ISGs (\SUPPFIG{ISG}).
}
\label{fig:transcriptomics}

@@

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were viral; this low-level background is likely from lysed cells that release ambient viral mRNA that is captured in droplets during reverse-transcription.
We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced too little mRNA to be detected in our experiment.
The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

We called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
We called presence / absence of genes rather than transcripts, since the two influenza genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
\FIG{transcriptomics}D (top panel) shows that 162 of 290 infected cells express all eight genes (see \SUPPFIG{frac_has_gene} for frequencies for individual genes).
This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work~\cite{russell2018extreme} and studies by others~\cite{brooke2013most, heldt2015single, dou2017analysis}, which have estimated that only 13\% to 50\% of infected cells express all genes.

@@

We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
Over 20\% of infected cells were IFN+ given the heuristic thresholds in \FIG{transcriptomics}G, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency.
Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
Many more cells expressed ISGs than IFN itself: the ratio of ISG+ to IFN+ cells was 1.8 among infected cells, and 7 among uninfected cells (\SUPPFIG{ISG}A).
The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in cells that do not directly detect infection~\cite{stetson2006type,honda2006type}.

@@

\subsection{Full genotypes of viruses infecting single IFN+ and IFN- cells}

%%% start genotypes figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[height=0.85\textheight]{figures/single_cell_figures/p_genotypes.pdf}
}
\caption{
Viral genotypes and infection outcomes in single cells.
Green and orange boxes at the left show the percent of all mRNA in that cell derived from virus and the percent of all cellular mRNA derived from IFN, respectively.
The second box is framed in orange for cells classified as IFN+ in \FIG{transcriptomics}G.
Blue arrows indicate the presence of a viral gene from one (light blue) or both (dark blue) viral barcode variants; a dark blue arrow therefore means that a cell was co-infected.
Circles and boxes on the arrows indicate mutations or indels as described in the legend at right.
The circle areas and box heights are proportional to the fraction of CCSs with that mutation.
For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively.
For instance, \textit{cell 5} was co-infected by a virion with one wild-type and one internally deleted copy of PB1.
}
\label{fig:genotypes}

@@

We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes expressed in single infected cells.
Using PCR enrichment (see Methods), we obtained over 200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
Crucially, the synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{StrandExchange}), meaning that the vast majority of CCSs correctly link the sequence of the viral transcript to cell barcodes and UMIs that identify the cell and molecule of origin.

After calling the presence / absence of each viral gene in each cell using the transcriptomic data as described in the previous section, we called mutations if they were found in at least two CCSs originating from different mRNAs (unique UMIs) and at least 30\% of all CCSs for that gene in that cell.
For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
This strategy reliably identifies mutations in virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
It will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
Therefore, analogous to somatic variant calling in tumor sequencing~\cite{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of transcripts in a cell.

@@

The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
Simple visual inspection of this figure reveals a wealth of information.
For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all eight genes and did not induce detectable IFN.
But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells expressed IFN.
As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) both lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
Many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).

However, \FIG{genotypes} also reveals stochasticity that is independent of viral genotype.
This stochasticity sometimes acts to the detriment of the virus, and sometimes to the detriment of the cell.
As an example of the former case, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the unmutated virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the unmutated virion that infected \textit{cell 105} still induced IFN.
But in other cases, the stochasticity allows a clearly defective virus to still escape immune recognition.
For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that do not activate IFN despite being infected by virions that fail to express NS.

\subsection{Viral defects associated with infection outcome in single cells}

>>>>

\newfloat{suppfile}{thp}{lofsupfile}
\floatname{suppfile}{\bf FILE}
\renewcommand{\thesuppfile}{S\arabic{suppfile}}

\newfloat{suppfig}{thp}{lofsupfile}
\floatname{suppfig}{\bf FIG}
\renewcommand{\thesuppfig}{S\arabic{suppfig}}

@@

% Allow big figures on pages
\renewcommand{\topfraction}{0.9}
\renewcommand{\bottomfraction}{0.9}
\renewcommand{\textfraction}{0.1}
\renewcommand{\floatpagefraction}{0.9}

@@

Unfortunately for the host, influenza initially only rarely triggers IFN production by infected cells~\cite{kallfass2013visualizing, killip2017single}.
This rareness of IFN induction is just one form of the extreme cell-to-cell heterogeneity that characterizes infection: cells also vary widely in their production of viral mRNA, proteins, and progeny virions~\cite{russell2018extreme,steuerman2018dissection,sjaastad2018distinct,heldt2015single,wang2018cell}.
Because viral growth and the IFN response are both feed-forward processes, early cell-to-cell heterogeneity could have significant downstream consequences for the race between virus and immune system---especially since natural human infections are typically initiated by just a few virions entering a few cells~\cite{mccrone2018stochastic, xue2018reconciling, varble2014influenza}.

It is unclear why only some infected cells trigger innate-immune responses.
Two possible contributors are pure stochasticity and pre-existing variation in cellular state.
For instance, only some cells induce IFN even upon treatment with synthetic innate-immune ligands~\cite{shalek2013single, shalek2014single, wimmers2018single}, and the frequency of IFN induction may depend on a cell's pre-existing chromatin state~\cite{bhushal2017cell}.
But for influenza, a third possible contributor also looms large: viral genetic diversity.
Because influenza has a high mutation rate, individual virions often have genetic defects~\cite{parvin1986measurement, suarez1992heterogeneity, suarez1994estimation, bloom2014experimentally, pauly2017novel}.
Indeed, many studies have identified mutations that increase IFN induction when engineered into a viral population~\cite{killip2017single, velthuis2018mini, du2018genome, perez2014unbiased}, and viral stocks that are rich in internal deletions in the polymerase genes induce more IFN~\cite{wang2018cell, baum2010preference, tapia2013defective, boergeling2015evidence, dimmock2015cloned}.

However, existing techniques are inadequate to determine how viral genetic diversity contributes to cell-to-cell heterogeneity during infection.
Flow cytometry and fluorescent reporters only measure protein levels~\cite{brooke2013most, guo2017single}, and current single-cell transcriptomic techniques measure abundance of transcripts and provide only fragmentary information on their sequences~\cite{russell2018extreme, steuerman2018dissection, wang2018cell, zanini2018single, zanini2018virus, saikia2018simultaneous, oneal2018west}.
None of these techniques reliably reveal if the virion infecting a specific cell has some idiosyncratic mutation.

Here we develop a new approach measure both the full transcriptome and sequences of all viral genes in single influenza-infected cells.
To do this, we perform both standard Illumina-based transcriptomics and full-length PacBio sequencing of viral genes from single cells.
Two-thirds of cells are infected by virions that have a mutation or defect in gene expression.
We identify several types of viral defects that increase IFN induction, but also show that viral diversity is insufficient to fully explain cell-to-cell heterogeneity during influenza infection.

@@

\subsection{A system to identify and enrich rare IFN+ cells.}

@@

Influenza virus only rarely triggers IFN expression in infected cells~\cite{killip2017single, russell2018extreme, kallfass2013visualizing}---a fact that poses a challenge for the study IFN induction in single cells.
Therefore, we developed a method to identify and enrich rare IFN+ cells by creating A549 cells that carried IFN reporters consisting of a type I (\textit{IFNB1}) or type III (\textit{IFNL1}) promoter driving expression of a cell-surface protein (LNGFR$\Delta$C~\cite{bonini1997hsv,ruggieri1997cell}) followed by a fluorescent protein (\FIG{IFNrare}A).
Cells that activate the IFN reporters can be enriched by magnetic-activated cell sorting (MACS) or identified by flow cytometry.
The reporters were efficiently activated by infection with a strain of Sendai virus~\cite{strahle2006sendai} that potently induces IFN (\SUPPFIG{reportervalidation}A), and activation of the type I and type III IFN reporters was highly correlated in our cells (\SUPPFIG{reportervalidation}B; further validated by the single-cell transcriptomics below).
Therefore, for the rest of this paper, we use ``IFN expression'' to refer to combined expression of type I and III IFNs.

We generated a stock of A/WSN/1933 (H1N1) influenza (hereafter referred to as ``WSN''), and found that it activated the IFN reporter in $\sim$0.5\% of infected cells  (\FIG{IFNrare}B), a frequency roughly comparable to that reported in prior studies~\cite{killip2017single, russell2018extreme}.
We also validated that MACS for the cell-surface protein driven by the IFN reporter enriched the IFN+ cells by $>$50-fold (\SUPPFIG{MACS}).

\subsection{Combined transcriptomics and virus-sequencing of single infected cells.}

@@

\caption{
Approach for combined transcriptomics and viral sequencing of single influenza-infected cells that express IFN.
(A)
IFN reporter A549 cells are infected with a mix of wild-type and synonymously barcoded viruses.
IFN+ cells are enriched by MACS, and pooled with non-enriched cells and uninfected canine cells that serve as a control for multiplets and mRNA leakage.
(B)
The mRNAs from individual cells are converted to cDNAs tagged with cell-specific barcodes.
(C)
Cellular transcriptomes are quantified using standard single-cell 3'-end Illumina sequencing, and 
(D)
viral genes are enriched by influenza-specific PCR and fully sequenced by PacBio (in this schematic, only the cell labeled by the red barcode is infected and has viral transcripts that are sequenced by PacBio).
(E)
The result is a matrix giving the expression of each gene in each cell, as well as the full sequences of the viral genes in infected cells.
}
\label{fig:workflow}

@@

We developed the approach in \FIG{workflow} to obtain the entire transcriptome \emph{and} the full sequences of all viral genes in single cells.
First, we generated a stock of virus that consisted of a mix of wild-type WSN and a ``synonymously barcoded'' variant that contained two engineered synonymous mutations near each termini of each gene (\SUPPFILE{virus_seqs}).
These viral barcodes allow us to identify co-infections from single-cell transcriptomic data~\cite{russell2018extreme}, and provide a control for PCR artifacts during full-length sequencing of viral transcripts (see below).
We used this viral stock to infect A549 IFN reporter cells (\FIG{workflow}A) at a dose that led to detectable viral transcription in $\sim$25\% of cells.
From 12 to 13 hours post-infection, we used MACS to enrich cells that activated the IFN reporter.
To ensure the presence of IFN- cells, we added back non-enriched cells to $\sim$10\% of the total.
We also added uninfected canine cells to $\sim$5\% of the total as a control for multiplets and to estimate the background amount of viral mRNA detected in truly uninfected cells.

We processed the cells on a commercially available platform~\cite{zheng2017massively} that isolates cells in droplets and reverse transcribes polyadenylated mRNAs to append a unique cell barcode to all cDNAs in each droplet, and a unique molecular identifier (UMI) to each cDNA molecule (\FIG{workflow}B).
Because influenza virus mRNAs are polyadenylated~\cite{robertson1981polyadenylation}, this process appends cell barcodes to viral as well as cellular mRNAs.
Furthermore, because virtually the entire influenza genome is transcribed, the cell-barcoded cDNA spans almost all 13,581 nucleotides in the segmented viral genome: the only portions not covered are one universally conserved nucleotide upstream of the transcription start site~\cite{koppstein2015sequencing} and 17 to 22 highly conserved nucleotides downstream of the polyadenylation site~\cite{robertson1981polyadenylation} in each of the eight viral gene segments.

We used a portion of the cell-barcoded cDNA for standard single-cell transcriptomics by Illumina 3'-end sequencing (\FIG{workflow}C).
But we also took a portion and enriched for full-length viral molecules by PCR (\FIG{workflow}D).
We performed PacBio sequencing on these full-length viral cDNAs to generate high-accuracy circular consensus sequences (CCSs)~\cite{travers2010flexible}.
These CCSs retain the cell barcodes, and with sufficient sequencing depth we obtain CCSs from multiple unique UMI-tagged cDNAs for each viral gene in each cell.
Because most cells are infected by just one or two virions, we can build a consensus of CCSs for each viral gene in each cell to determine the sequence(s) of these virions.
Combining this information with the 3'-end sequencing determines the entire transcriptome and the full sequences of the infecting virions in single cells (\FIG{workflow}E).

\subsection{Transcriptomic analyses of single IFN+ and IFN- influenza-infected cells.}

@@

\caption{
Single-cell transcriptomics of IFN-enriched influenza-infected cells.
(A) 
Number of cells for which transcriptomes were obtained.
From these numbers, we estimate~\cite{bloom2018estimating} that $\approx$11\% of the transcriptomes are derived from multiple cells. 
(B) The number of cellular and viral mRNAs detected for each cell is plotted as a point.
Green lines show the distribution of cellular mRNAs per cell.
Cells outside the dashed magenta lines have unusually low or high amounts of cellular mRNA (likely low-quality emulsions or multiplets), and are excluded from subsequent analyses.
(C) Distribution across cells of the fraction of all mRNA derived from influenza.
Cells called as infected are in blue, while other cells are in green.
The inset shows the amount of viral mRNA in the human cells that are called as infected.
(D) Number of influenza genes detected per infected cell, and the amount of viral mRNA in cells expressing each number of viral genes.
\SUPPFIG{frac_has_gene} shows the frequency that each viral gene is detected.
(E) Relative expression of viral genes, quantified as the fraction of all viral mRNA in each infected cell derived from each gene.
(F) Number of cells infected with wild-type virus, synonymously barcoded virus, or both.
From the cells infected with both viral barcodes, we estimate~\cite{bloom2018estimating} that 63\% of infected cells are co-infected.
(G) Fraction of cellular mRNA from IFN across cells, faceted by whether the cells are infected.
Cells to the left of the first dashed magenta line are classified as IFN-, and cells to the right of the second line as IFN+.
A pseudocount is added to the number of IFN transcripts detected in each cell, which is why none of the fractions are zero.
}
\label{fig:transcriptomics}

@@

To identify infected cells, we examined the fraction of each transcriptome derived from virus (\FIG{transcriptomics}C).
As expected, only a small fraction ($\sim$0.7\%) of transcripts in the uninfected canine cells were viral; this low-level background is likely from lysed cells that release ambient viral mRNA.
We tested whether each cell contained significantly more viral transcripts than expected under a Poisson model given this background fraction, and classified 290 human cells as definitively infected with influenza (\FIG{transcriptomics}C).
We classified the other cells as uninfected, although it is possible that some were infected with virions that produced very little mRNA.
The distribution of the amount of viral mRNA across infected cells is shown in the inset in \FIG{transcriptomics}C.
As in our prior work~\cite{russell2018extreme}, the distribution is extremely heterogeneous: many infected cells have only a few percent of mRNA derived from virus, but viral mRNA comprises over half the transcriptome of a few cells.

We called the presence or absence of each viral gene in each infected cell, again using a Poisson model parameterized by background fractions estimated from uninfected canine cells.
We called presence / absence of genes rather than transcripts, since the two genes that encode multiple transcripts (M1 / M2 from the M gene, and NS1 / NS2 from the NS gene) do so via alternative splicing that leaves both isoforms with the same termini, making them indistinguishable by 3'-end sequencing.
\FIG{transcriptomics}D (top panel) shows that 162 of 290 infected cells express all eight genes (see \SUPPFIG{frac_has_gene} for frequencies for individual genes).
This measured frequency of infected cells expressing all eight genes is slightly higher than in our own prior work using the WSN strain~\cite{russell2018extreme}, and slightly to substantially higher than that reported in studies by others using different viral strains or methodologies~\cite{heldt2015single, brooke2013most, dou2017analysis, jacobs2019incomplete}.

@@

We next examined expression of IFN and ISGs (\FIG{transcriptomics}G and \SUPPFIG{ISG}).
Over 20\% of infected cells were IFN+, indicating that the MACS successfully enriched IFN+ cells far beyond their initial frequency.
Few ($\sim$1.3\%) uninfected cells were IFN+; the few that were present might be because the MACS enriched for rare cells that spontaneously activated IFN, or because some cells that we classified as uninfected were actually infected at low levels.
Many more cells expressed ISGs than IFN itself (\SUPPFIG{ISG}A).
The IFN+ cells were a subset of the ISG+ cells: IFN+ cells always expressed ISGs, but many ISG+ cells did not express IFN (\SUPPFIG{ISG}B).
These results are consistent with the established knowledge that IFN is expressed only in cells that directly detect infection, but that ISGs are also expressed via paracrine signaling in other cells~\cite{stetson2006type,honda2006type}.

@@

\subsection{Full genotypes of viruses infecting single IFN+ and IFN- cells.}

%%% start genotypes figure %%%
\begin{figure}
\begin{fullwidth}
{\centering
\includegraphics[height=0.85\textheight]{figures/single_cell_figures/p_genotypes.pdf}
}
\caption{
Viral genotypes and infection outcomes in single cells.
Green and orange boxes at the left show the percent of all mRNA in that cell derived from virus and the percent of all cellular mRNA derived from IFN, respectively.
The second box is framed in orange for cells classified as IFN+ in \FIG{transcriptomics}G.
Blue arrows indicate the presence of a viral gene from one (light blue) or both (dark blue) viral barcode variants; a dark blue arrow therefore means that a cell was co-infected.
Circles and boxes on the arrows indicate mutations or indels as described in the legend at right.
The circle areas and box heights are proportional to the fraction of CCSs with that mutation.
For dual-barcode infections, mutations / indels for the wild-type and synonymously barcoded viral variants are shown in the top and bottom half of the arrows, respectively.
For instance, \textit{cell 5} was co-infected by a virion with one unmutated and one internally deleted copy of PB1.
}
\label{fig:genotypes}

@@

We next used PacBio sequencing (\FIG{workflow}D) to determine the full sequences of the viral genes in single infected cells.
We obtained $>$200,000 high-quality PacBio CCSs that mapped to an influenza gene and contained a cell barcode and UMI (\SUPPFIG{CCSs}).
The synonymous viral barcodes at both termini of each gene enabled us to confirm that PCR strand exchange was rare (\SUPPFIG{StrandExchange}), meaning that the vast majority of CCSs correctly link the sequence of the transcript to cell barcodes and UMIs that identify the cell and molecule of origin.

After calling the presence / absence of each viral gene in each cell as described in the previous section, we called mutations if they were found in at least two CCSs originating from different mRNAs (unique UMIs) and at least 30\% of all CCSs for that gene in that cell.
For cells co-infected with both viral barcode variants, we called mutations separately for each viral variant.
This strategy reliably identifies mutations in virions that initiate infection of cells infected with at most one virion of each viral barcode variant ($\sim$75\% of infected cells), as well as high-abundance mutations in cells co-infected with multiple virions of the same viral barcode.
It will not identify mutations that arise within a cell after the first few rounds of viral genome replication, since such mutations will not reach 30\% frequency in that cell.
Therefore, analogous to somatic variant calling in tumor sequencing~\cite{xu2014comparison, cibulskis2013sensitive}, there is a limit to our detection threshold: we cannot identify mutations that occur on just a small fraction of transcripts in a cell.

@@

The 150 cells for which we called the full viral genotypes are shown in \FIG{genotypes}.
Visual inspection of this figure reveals a wealth of information.
For instance, the cell with the highest viral burden (\textit{cell 1} in \FIG{genotypes}, which has 65\% of its mRNA from virus) was infected by a virion that expressed unmutated copies all eight genes and did not induce detectable IFN.
But 12 of the other 13 cells with at least 50\% of their mRNA from virus were infected by virions that had a mutation or failed to express a gene, and five of these cells produced IFN.
As expected, all cells infected by virions that failed to express a component of the viral polymerase complex (PB2, PB1, PA, or NP) expressed low amounts of viral mRNA since they are limited to primary transcription off the incoming proteins (e.g., \textit{cell 132} and \textit{cell 143}).
The two cells that expressed the most IFN (\textit{cell 13} and \textit{cell 123}) lacked the viral NS gene that encodes the virus's primary IFN antagonist~\cite{garcia1998influenza, hale2008multifunctional}.
Many other IFN+ cells had different defects such as large internal deletions (e.g., \textit{cell 5} and \textit{cell 89}) or amino-acid mutations (e.g., \textit{cell 9}, \textit{cell 28}, and many others).

However, \FIG{genotypes} also reveals stochasticity that is independent of viral genotype.
This stochasticity sometimes acts to the detriment of the virus, and sometimes to the detriment of the cell.
As an example of the former case, expressing unmutated copies of all eight genes did not guarantee a favorable outcome for the virus: for instance, the unmutated virion that infected \textit{cell 139} only managed to express viral mRNA to 6\% of the total transcriptome, and the unmutated virion that infected \textit{cell 105} still induced IFN.
But in other cases, the stochasticity allows a defective virus to still escape immune recognition.
For instance, there are a number of cells (e.g., \textit{cell 62} and \textit{cell 78}) that do not activate IFN despite being infected by virions that fail to express NS.

\subsection{Viral defects associated with infection outcome in single cells.}
